RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data

### 2015-02-23 doc/bioinformatics2014/revision2/response_to_reviews.txt

3. Instead of using DNA sequencing of primary tumors genomes, which are highly heterogeneous and present many rare (single nucleotide) variants, the authors used the DNA from a commercially available cell line, which is colonially expanded (genetically homogenous). Interestingly, chromosome 7, where the investigated gene is located, is tetraploid on HCC1187[1], therefore, one would expect to find only allele frequencies 0%, 25%, 50%, 75 and 100% on variants located in this chromosome. Figure 5 seems to be in agreement to this hypothesis, except for variants 49704 and 81769, which has marginally lower estimations of allele frequency.Therefore, the variants detected on HCC1187 are not actually rare variants (MAF=25%) and, as shown on Figure 3 (Table1), may be detected by most of currently available methods. Despite the limitations, the analysis is complete, but it would be more informative to apply RVD2 to publicly available (and validated) DNAseq from primary tumors.

4. The authors cite a paper from 2011 to refer for the Synthetic DNA sequencing. Explicit details on the technology, read length and number of reads generated would greatly ease readers understanding of this section.

Section 4.1 describing the details of the experimental protocol for the synthetic sequencing data is indeed brief. We have added the details the reviewer suggested and as expected, the narrative is much clearer. Briefly, Supplementary Table S1 in the Nucleic Acid Research publication shows that each sample has roughly 1,000,000 35bp paired-end reads with no errors in the multiplexing barcodes to give an average read depth of ~100,000 across the synthetic amplicon. We describe this and more detail as suggested in the protocol section.

5. The authors greatly expose the comparison of sensibility, specificity, true positive and true negative rates. It seems that the caveat of the method is on memory consumption and
processing time. It would be very informative to explicitly have a table with execution time
and memory consumption of all utilized methods, preferably in a more realistic scenario,
more specifically, a whole genome analysis.

The memory consumption and computational time for the algorithm is an important consideration. We intentionally focused on the accuracy of the methodology here and so chose Metropolis-Hastings sampling as our inference algorithm. Metropolis-Hastings is known to converge to sampling from the true posterior distribution in the limit. As such, presently the algorithm is not as computationally fast as it could be. To improve the speed of the algorithm, we have implemented a multi-threaded version to take advantage of a multi-core computational configuration. We provide computational speed results in Section X.X and discuss avenues for improvement that we are currently pursuing.

### 2015-02-25 doc/bioinformatics2014/revision2/response_to_reviews.txt

The reviewer correctly points out that much next-generation DNA sequencing data has a median depth in the 30x - 100x range. We have added a paragraph to our discussion of the sensitivity and specificity results (second paragraph in Section 5.2). In that paragraph, we discuss three primary median depth ranges and typical applications for those areas. Briefly, whole genome sequencing typically has depths in the range of 10x - 100x as noted. In addition, targeted cancer gene sequencing typically has median read depths in the range 100x - 1,000x because the region of interest is smaller and there is interest in measuring tumor heterogeneity. Microbial and viral sequencing can have median depths from 1,000x - 100,000x because one is interested in identifying rare subpopulations that may prove resistant to antibiotics or antivirals.

We have chosen a broad range of median depths for our results so that researchers in whole genome sequencing, targeted sequencing and microbial sequencing can see how the methodology would perform in their application area.

2. Despite having the whole genome sequencing for HCC1187 and HCC1187BL cell lines, the authors focus all presented analysis one a single gene. It's not clear why the authors used the coordinates chr7:154738059154782774 (~45Kb), excluding the three first exons (four introns) and the last exon and most of the last intron, when, in fact, PAXIP1 has ~59Kb. No further explanation is given regarding why this specific region was chosen.

Thank you for identifying this important omission. The region of interest for the PAXIP1 gene should have covered the entire 59Kb coding region. We ran RVD2 on the current NCBI start/stop codons (http://www.ncbi.nlm.nih.gov/gene/22976) to give the expanded region of chr7:154,735,400 - 154,794,682. This expanded region yields additional mutations. In particular, using the expanded region, we identified 5 additional germline variants and 2 additional somatic variants. In expanding the region of interest the previously identified germline variants at chr7:154,766,700 and chr7:154,781,769 are not called. A close inspection shows that these loci have very similar point estimates of the allele frequency in the case and control in the shortened region of interest. In the analysis with the expanded region, RVD2 estimates a lower across-locus precision parameter (M0) and thus does not call these loci. The control M0 parameter is 5.7 in the previous analysis and 4.2 in the 59kbp region.

### 2015-02-25 doc/bioinformatics2014/revision2/document.tex
Updated narrative for PAXIP1 region.

\vspace{-10pt}
\subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}
\begin{figure*}[!hbpt]
\centering
\vspace{-10pt}
\includegraphics[width=0.68\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
\caption{Estimated minor allele fraction for germline and somatic mutations called by RVD2 in the 44kbp PAXIP1 gene. Blue diamonds ($\diamond$) indicates germline mutations, where $\mu^{\text{control}}$ is significantly different from the reference sequence. Red stars (*) indicates somatic mutations, where $\mu^{\text{case}}$ is significantly different from $ \mu^{\text{control}}$. The  vertical lines represent 95\% credible interval around posterior mean MAF. Five positions are common population SNPs according to dbSNPv138, and the identities are shown below the positions.}
\vspace{-10pt}
\label{fig:brca_MAF}
\end{figure*}

\begin{figure*}[!pthb]
\vspace{-10pt}
    \centering
    \subfloat[\label{fig:brca_dot:a}]{{\includegraphics[width=0.7\textwidth]{pdf_figs/HCC1187_dotplot.pdf} }}%
    %
    \subfloat[\label{fig:brca_dot:b}]{{\includegraphics[width=0.18\textwidth]{pdf_figs/position_lookup_chart.pdf} }}%
    \caption{(a) Positions called by VarScan2-somatic, muTect, RVD2 and Strelka in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774. The positions are sorted by index (correspondance to genomic positions shown in Supplementary Table 1) (b) Read counts for each base for positions called by RVD2 and muTect from raw pileup data.}%
    \label{fig:brca_dot}%
\vspace{-10pt}
\end{figure*}

RVD2 identified twelve variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774. There were eight germline variants and eight somatic mutations. %Especially, two heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. Positions chr7:154743899C\textgreater T, chr7:154749704G\textgreater A, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154766700C\textgreater A, chr7:154780960C\textgreater T, and chr7:154781769G\textgreater T were called germline mutations. Positions chr7:154749704G\textgreater G, chr7:154753635T\textgreater C, \\ chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154760439A\textgreater C, chr7:154766732T\textgreater G,  chr7:154766832A\textgreater C, chr7:154777118A\textgreater C were identified as significantly different in tumor and normal sample MAF and called somatic mutation. Positions chr7:154754371 and chr7:154758813 appear to be loss-of-heterozygosity events. Some of these mutations are also found to be common population SNPs according to dbSNPv138. The corresponding identities are shown in the Figure~\ref{fig:brca_MAF}. The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1. 
%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.

>>>>

\vspace{-10pt}
\subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}
\begin{figure*}[!hbpt]
\centering
\vspace{-10pt}
\includegraphics[width=0.68\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
\caption{Estimated minor allele fraction for germline and somatic mutations called by RVD2 in the 44kbp PAXIP1 gene. Blue diamonds ($\diamond$) indicates germline mutations, where $\mu^{\text{control}}$ is significantly different from the reference sequence. Red stars (*) indicates somatic mutations, where $\mu^{\text{case}}$ is significantly different from $ \mu^{\text{control}}$. The  vertical lines represent 95\% credible interval around posterior mean MAF. Five positions are common population SNPs according to dbSNPv138, and the identities are shown below the positions.}
\vspace{-10pt}
\label{fig:brca_MAF}
\end{figure*}
% figure 5 generated from 2014-03-24_germline_and_somatic_test_correction_no_subclassification

\begin{figure*}[!pthb]
\vspace{-10pt}
    \centering
    \subfloat[\label{fig:brca_dot:a}]{{\includegraphics[width=0.7\textwidth]{pdf_figs/HCC1187_dotplot.pdf} }}%
    %
    \subfloat[\label{fig:brca_dot:b}]{{\includegraphics[width=0.18\textwidth]{pdf_figs/position_lookup_chart.pdf} }}%
    \caption{(a) Positions called by VarScan2-somatic, muTect, RVD2 and Strelka in the 44kbp PAXIP1 gene from chr7:154,735,400 to chr7:154,794,682. The positions are sorted by index (correspondance to genomic positions shown in Supplementary Table 1) (b) Read counts for each base for positions called by RVD2 and muTect from raw pileup data.}%
    \label{fig:brca_dot}%
\vspace{-10pt}
\end{figure*}

% figure 6 generated from 2014-03-12_dot_plot_mutation_detection_result_HCC1187_dataset

RVD2 identified twelve variants in the 59kbp PAXIP1 gene from chr7:154,735,400 to chr7:154,794,682. There were 11 germline variants and 10 somatic mutations. %Especially, two heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. 
Positions 
chr7:154,743,899C\textgreater T, 
chr7:154,749,704G\textgreater A, 
chr7:154,753,635T\textgreater C, 
chr7:154,754,371T\textgreater C, 
chr7:154,758,813G\textgreater A, 
chr7:154,780,960C\textgreater T, 
chr7:154,784,369A\textgreater G, 
chr7:154,787,357C\textgreater T, 
chr7:154,787,367C\textgreater T, 
chr7:154,787,612A\textgreater G, and 
chr7:154,788,798A\textgreater G 
were called germline mutations. 
%
Positions 
chr7:154,749,704G\textgreater A, 
chr7:154,753,635T\textgreater C, 
chr7:154,754,371T\textgreater C, 
chr7:154,758,813G\textgreater A, 
chr7:154,760,439A\textgreater C, 
chr7:154,766,732T\textgreater G,  
chr7:154,766,832A\textgreater C,
chr7:154,777,118A\textgreater C,
chr7:154,787,612A\textgreater C,
chr7:154,788,798A\textgreater C, and
were identified as significantly different in tumor and normal sample and called somatic mutations.
% 
Positions chr7:154,754,371 and chr7:154,758,813 appear to be loss-of-heterozygosity events. 
Some of these mutations are also found to be common population SNPs according to dbSNPv138. 
The corresponding identities are shown in the Figure~\ref{fig:brca_MAF}. The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1. 
%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
